Annotation exploration for SCPCP000006

### 2024-10-21 analyses/cell-type-wilms-tumor-06/notebook/04_annotation_Across_Samples_exploration.Rmd
Add files via upload

---
title: "Annotation exploration for SCPCP000006"
author: "Maud PLASCHKA"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: yes
    toc_float: yes
    code_folding: hide
    highlight: pygments
    df_print: paged
---

@@

      # Create and return a data frame from the Seurat object with relevant annotations
      # this data frame will have four columns: barcode, sample_id, compartment, organ
      data.frame(
        compartment = srat$fetal_kidney_predicted.compartment,
        organ = srat$fetal_full_predicted.organ, 
        PTPRC = FetchData(object = srat, vars = "ENSG00000081237", layer = "counts"),
        VWF = FetchData(object = srat, vars = "ENSG00000110799", layer = "counts")
        ) |>
        tibble::rownames_to_column("barcode") |>
        dplyr::mutate(sample_id = sample_id)
    }
  ) |>
  # now combine all dataframes to make one big one
  dplyr::bind_rows()

@@

@@

@@

>>>>

---
title: "Annotation exploration for SCPCP000006"
author: "Maud PLASCHKA"
date: "`r Sys.Date()`"
params: 
  mapping_score_thr: 0.85
output: 
  html_document: 
    toc: yes
    toc_float: yes
    code_folding: hide
    highlight: pygments
    df_print: paged
---

@@

      # Create and return a data frame from the Seurat object with relevant annotations
      # this data frame will have four columns: barcode, sample_id, compartment, organ
      data.frame(
        compartment = srat$fetal_kidney_predicted.compartment,
        compartment_score = srat$fetal_kidney_predicted.compartment.score,
        organ = srat$fetal_full_predicted.organ, 
        PTPRC = FetchData(object = srat, vars = "ENSG00000081237", layer = "counts"),
        VWF = FetchData(object = srat, vars = "ENSG00000110799", layer = "counts")
        ) |>
        tibble::rownames_to_column("barcode") |>
        dplyr::mutate(sample_id = sample_id)
    }
  ) |>
  # now combine all dataframes to make one big one
  dplyr::bind_rows()

cell_type_df <- merge(cell_type_df, metadata, by.x = "sample_id", by.y = "scpca_sample_id")

cell_type_df$pass_mapping_QC <- FALSE 
cell_type_df$pass_mapping_QC[cell_type_df$compartment_score > params$mapping_score_thr] <- TRUE
```

@@

- `df` is the name of the table containing metadata and feature expression (counts) per cells
- `feature` is the name of the gene to average the expression
- `group.by` is the name of the metadata to group the cells
- `split.by` is the name of the metadata to split the plots

@@

#### do_Feature_densityplot

`do_Feature_densityplot` shows boxplot of expression of a feature grouped by a metadata.

```{r fig.width=10, fig.height=4, out.width='100%'}
do_Feature_densityplot <- function(df, group.by, feature, split.by = NULL  ){

  p <- ggplot(df, 
    aes(x = !!sym(feature),  fill = group)
  ) + 
  geom_density() +
  facet_wrap(vars(!!sym(group.by)), scale = "free_y", ncol = 4) + 
  theme_bw()

@@

DT::datatable(compartment_df, 
              caption = "counts of cells in each compartment",
              extensions = 'Buttons', 
              options = list(  dom = 'Bfrtip',
                               buttons = c( 'csv', 'excel')))
```
### Label transfer (mapping) scores for the four compartments

```{r fig.height=5, fig.width=10, message=FALSE, warning=FALSE, out.width='100%'}
p <- do_Feature_densityplot( df = cell_type_df,
                    feature = "compartment_score",
                    group.by = "compartment"
                    )
p +  geom_vline(xintercept=params$mapping_score_thr)
```

Some cells have an excellent mapping score (>0,85) while other performed worth. Let's have a look at the repartiction per patient.

```{r fig.height=15, fig.width=15, message=FALSE, warning=FALSE, out.width='100%'}
p <- do_Feature_boxplot( df = cell_type_df,
                    feature = "compartment_score",
                  group.by = "compartment",
                    split.by = "sample_id")

p + geom_hline(yintercept = params$mapping_score_thr)

Regarding the endothelial cells, some patient have very high mapping scores while others perform poorly. 
Let's have a look if this can be linked to some metadata/clinical data.

##### Treatment

```{r fig.height=3, fig.width=10, message=FALSE, warning=FALSE, out.width='100%'}
p <- do_Feature_boxplot( df = cell_type_df,
                    feature = "compartment_score",
                  group.by = "compartment",
                    split.by = "treatment")

p + geom_hline(yintercept = params$mapping_score_thr)

##### Subdiagnosis

```{r fig.height=3, fig.width=10, message=FALSE, warning=FALSE, out.width='100%'}
p <- do_Feature_boxplot( df = cell_type_df,
                    feature = "compartment_score",
                  group.by = "compartment",
                    split.by = "subdiagnosis")

p + geom_hline(yintercept = params$mapping_score_thr)

##### Disease timing

```{r fig.height=3, fig.width=15, message=FALSE, warning=FALSE, out.width='100%'}
p <- do_Feature_boxplot( df = cell_type_df,
                    feature = "compartment_score",
                  group.by = "compartment",
                    split.by = "disease_timing")

p + geom_hline(yintercept = params$mapping_score_thr)

We do not really see a pattern linking the label transfer score and some meta/clinical data. 
For the later CNV analysis, we can apply a threshold of `r params$mapping_score_thr` for the `fetal_kidney_predicted.compartment.score` while selecting for normal cells, i.e. `immune` and/or `endothelial` cells.

We finally look for each patient and compartment, the number of cells that have a mapping score > `r params$mapping_score_thr` (TRUE).

```{r fig.width=25, fig.height=3, results='asis'}
scores_df <- cell_type_df |>
  # count how many per compartment
  dplyr::count(sample_id, compartment, pass_mapping_QC) |>
  # make the table wide and use a value of 0 if a count is missing
  tidyr::pivot_wider(names_from = pass_mapping_QC, values_from = n, values_fill = 0)

DT::datatable(scores_df, 
              caption = "counts of cells that have a good/poor mapping in each compartment",
              extensions = 'Buttons', 
              options = list(  dom = 'Bfrtip',
                               buttons = c( 'csv', 'excel')))
```
